Normalization by Evaluation for Sized Dependent Types

### 2017-02-24 nbe/icfp17.tex
[ ICFP 17 ] Refactored the logical relation.

In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.  The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.  However, the logical relation defined in the following will also yield a weak head normalization theorem.

\[
  \fbox{$\LRSize \Gamma a \alpha$} \defas
  \forall \xi : \Gamma' \leq \Gamma.\ \
  \RSize {\Gamma'} \alpha {a\xi}
  % \exists a'.\
  % \RSize {\Gamma'} \alpha a' \mand
  % \Gamma' \der a\xi = a' : \Size
\]

We simultaneously define
\fbox{$\LRT \Gamma {T'} {A'} \ell$} for $\Gamma \der T' : \Set[\ell]$
and \fbox{$\LR \Gamma t {T'} f {A'}$} for $\Gamma \der t : T'$ and $f \in A'$
defined by induction on ${A'} \in \SET[\ell]$.

@@

  \noindent
  $\LRT{\Gamma}{T'}{A'}\ell$ iff
  $\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\forall \xi : \Gamma' \leq \Gamma.\ \
   \exists m. \ \
   \R {\Gamma'} {\down {A'} f} m \mand
   \Gamma' \der t\xi = m : T'\xi$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and \\
  $\forall \xi : \Gamma' \leq \Gamma.\ \
   \exists v. \ \
   \R {\Gamma'} {\down {A'} f} v \mand
   \Gamma' \der t\xi = v : \Nat\,a\xi$.

@@

\begin{theorem}[Into and out of the logical relation]
  \label{thm:into}
  \bla
  Let $A \in \SET[\ell]$.
  \begin{enumerate}

  \item \label{it:into}
     $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
     $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.

  \item \label{it:outty}
     If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
     there is some normal form $V$ such that
     $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.

  \item \label{it:out}
    If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
    there is some normal form $v$ such that
    $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.

@@

\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
\begin{enumerate}
\item To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
Thus, it is sufficient to show that there is some normal form $v$ such that
$\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
However, this follows by induction hypothesis (\ref{it:out}) on
assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}
\end{caselist}
\end{proof}
\begin{corollary}[One-to-one]
\label{cor:oneone}
\bla
\begin{enumerate}
\item
If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
\item
If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
\end{enumerate}
\end{corollary}

>>>>

In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.  The proof follows \citet{abelCoquandDybjer:lics07,abel:habil}.  However, the logical relation defined in the following will also yield a weak head normalization theorem.
\begin{alignat*}{8}
& \Gamma \der a && \doteq \tRsize\, \alpha
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \mathrlap{ \RSize {\Gamma'} \alpha {a\xi} }
\\[-0.6ex]
& \Gamma \der T && \doteq \tRty\, A  : s
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists V.\
  && \RTy{\Gamma'} A V
  && ~\mand~ \Gamma' \der T\xi = V : s
\\[-0.6ex]
& \Gamma \der t && \doteq \tR\, d  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists v.\
  && \R{\Gamma'} d v
  && ~\mand~ \Gamma' \der t\xi = v : T\xi
\\[-0.6ex]
& \Gamma \der t && \doteq \tRne\, n  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists m.\
  && \RNe{\Gamma'} n m
  && ~\mand~ \Gamma' \der t\xi = m : T\xi
\end{alignat*}
% \[
% \begin{array}{lll}
% \Gamma \der a \doteq \tRsize\, \alpha
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \ \RSize {\Gamma'} \alpha {a\xi}
% \\
% \Gamma \der T \doteq \tRty\, A : s
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists V.\
%   \RTy{\Gamma'} A V \mand
%   \Gamma' \der T\xi = V : s
% \\
% \Gamma \der t \doteq \tR\, d : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists v.\
%   \R{\Gamma'} d v \mand
%   \Gamma' \der t\xi = v : T\xi
% \\
% \Gamma \der t \doteq \tRne\, n : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists m.\
%   \RNe{\Gamma'} n m \mand
%   \Gamma' \der t\xi = m : T\xi
% \end{array}
% \]
% \[
%   \fbox{$\LRSize \Gamma a \alpha$} \defas
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \RSize {\Gamma'} \alpha {a\xi}
%   % \exists a'.\
%   % \RSize {\Gamma'} \alpha a' \mand
%   % \Gamma' \der a\xi = a' : \Size
% \]
By definitions, these relations are closed under subsumption and
weakening, e.g., if $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
and if $\xi : \Gamma \leq \Gamma'$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.

\begin{lemma}[Closure properties for neutrals]
\label{lem:rclosne}
\bla
\begin{enumerate}
\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
\item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
\end{enumerate}
\end{lemma}

We simultaneously define
\fbox{$\LRTS \Gamma {T'} {A'} s$} for $\Gamma \der T' : s$
and \fbox{$\LR \Gamma t {T'} f {A'}$} for $\Gamma \der t : T'$ and $f \in A'$
by induction on ${A'} \in s$.

@@

  \noindent
  $\LRTS{\Gamma}{T'}{A'} s$ iff
  $\Gamma \der T' \doteq \tRty\,A' : s$
  %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists m. \ \
  %  \R {\Gamma'} {\down {A'} f} m \mand
  %  \Gamma' \der t\xi = m : T'\xi$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' = \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and
  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$
  % \\
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists v. \ \
  %  \R {\Gamma'} {\down {A'} f} v \mand
  %  \Gamma' \der t\xi = v : \Nat\,a\xi$.

@@

\begin{theorem}[Into and out of the logical relation]
\label{thm:into}
\bla
Let $A \in s$.
\begin{enumerate}

\item \label{it:into}
   If\/ $\Gamma \der t \doteq \tRne\,n : T$ then
   $\LR \Gamma t T {\up A n} A$.

\item \label{it:outty}
   If\/ $\LRTS{\Gamma} T A s$ then
   $\Gamma \der T \doteq \tRty\,A : s$

\item \label{it:out}
  If\/ $\LR \Gamma t T f A$ then
  $\Gamma \der t \doteq \tR\,\down A f : T$.

% \item \label{it:into}
%    $\LR \Gamma t T {\up A n} A$ if for all $\xi : \Gamma' \leq \Gamma$ there is an $m$ such that
%    $\RNe{\Gamma'}n m$ and\/ $\Gamma' \der t\xi = m : T$.

% \item \label{it:outty}
%    If\/ $\LRT{\Gamma} T A \ell$ then for all $\xi : \Gamma' \leq \Gamma$
%    there is some normal form $V$ such that
%    $\Rty{\Gamma'} A V$ and $\Gamma' \der T\xi = V : \Set[\ell]$.

% \item \label{it:out}
%   If\/ $\LR \Gamma t T f A$ then for all $\xi : \Gamma' \leq \Gamma$
%   there is some normal form $v$ such that
%   $\R{\Gamma'} {\down A f} v$ and $\Gamma' \der t\xi = v : T\xi$.

@@

\nextcase $A' \evalsto \piT A B$ and $\resurrect\Gamma \der T' = \piT U T$.
\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {A'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{B[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down A a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down A a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}
\end{caselist}
\end{proof}

\begin{corollary}[One-to-one]
\label{cor:oneone}
\bla
\begin{enumerate}
\item
If $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
\item
If $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
\end{enumerate}
\end{corollary}
